---
title: "Introduction to dplyr"
author: "Fabrice Rossi"
format: 
  revealjs:
    echo: false
  pptx:
    echo: false
lightbox: true
df-print: kable
---

```{r}
here::i_am("dplyr-101.Rproj")
library(here)
```

```{r}
library(ggplot2)
theme_set(theme_bw())
library(dplyr)
```

```{r}
data("diamonds")
```

# Introduction to filter

## Link between diamond carat and price

```{r}
ggplot(diamonds, aes(x=carat, y=price)) +
  geom_point()
```

## Selecting a subset of a dataset

```{r}
#| echo: true
diamonds_2ct <- diamonds |>
  filter( abs(carat - 2) < 0.1)
```

```{r}
ggplot(diamonds_2ct, aes(x=carat, y=price)) +
  geom_point()
```

## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x=carat,y=price, group=carat)) +
  geom_boxplot()
```

## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x=carat,y=price, group=carat)) +
  geom_violin()
```

##Using filter {.smaller} Filter is used as follows

```{r}
#| eval: false
#| echo: true
filter(dataframe, expression with a logical value)

```

code keeps the rows of the dataframe for which the logical is true

```{r}

filter( diamonds, price==1000) |>
  ggplot(aes(y=cut))+
  geom_bar()
```

## Multiple logical expressions

::::: (.columns)
::: {.column width="40%"}
Filter is used as follows

```{r}
#| eval: false
#| echo: true
filter(dataframe, exp1, exp2, exp3)

```

code keeps the rows of the dataframe for which all expressions are true
:::

::: {.column width="60%"}
```{r}

filter( diamonds, price==1000, cut=="Ideal") |>
  ggplot(aes(y=color))+
  geom_bar()
```
:::
:::::

# Logical Expressions

## comparison operators

Everything that compares two values and returns `TRUE`or `FALSE`.

```{r}
#| echo: true

x<- 2
x>=3
x<2
x==2


```

## Logical operators

combining true values

```{r}
#| echo: true
TRUE & FALSE
TRUE | FALSE
!FALSE
```

`&`is and `|`is or `!`is NOT

## Examples

equivalent filter expresions :

```{r}
#| echo: true
#| eval: false
diamonds |> filter( price==1000, cut=="Ideal")
diamonds |> filter((price==1000) & ( cut=="Ideal"))
 
```

different filter expressions:

```{r}
#| echo: true
#| eval: false

diamonds |> filter( price==1000, cut=="Ideal")
diamonds |> filter((price==1000) | ( cut=="Ideal"))
 
```

# Introduction to `summarise`

## Global Summaries

```{r}
#| echo: true
diamonds |> 
  summarise(mean(price), 
            median(price))
```

`summarise` is used as follows

```{r}
#| eval: false
#| echo: true
summarise(dataframe, expression based on aggregation function)
```

The result is a dataframe with one column per expression and one row.

## Aggregation functions {.smaller}

::::: columns
::: {.column width="50%"}

An aggregation function computes a statistics (in a broad sense) on a numerical vector for instance:

- `mean`
- `median`
- `sd` : Standard Deviation
- `var` : Variance
- `max` and `min`

:::

::: {.column width="50%"}

```{r}
#| echo: true

x<- c(0.1,1,2, 2, -3, 5)

mean(x)
median(x)
sd(x)
min(x)
n_distinct(x)
length(x)

```

:::
:::::
